\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2016-10-31 doc/publications/plos16/plos16.tex
changes to plos16.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Feature Selection on real data 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ selects previously demonstrated cancer variants}
%Feature selection can be useful for determining variables the contribute to the label or for perhaps selecting the n most important
%variables for further analysis. 
As established in the pervious section, \cursedforest\ allows the user to retrieve a subsection of features, ranked by
their importance to identify the variables predictive of the label or selecting the $n$ most important variables for
further analysis.  We demonstrate this feature on The Cancer Genome Atlas (TCGA) BRCA dataset.  We build this model on
VCF files retrieved from the TCGA data portal.  For the label, we use the cell type (i.e. tumor or normal).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 3: Classification selection
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ outperforms existing methods for classification problems}
As established in Section~\ref{synthetic}, \cursedforest\ 
In this section we compare \cursedforest 's performance against other published methods for predicting ethnicity on the
1000 Genomes Project dataset.

@@

\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.077 (XX) &                  &                                                                   \\
                                                & Spark ML  &            &                  &                                                                   \\
                                                & RF - R         & 0.05       & 24 min           & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes} \\
                                                & ranger - R      & 0.94       & 1 min 10 seconds & 16GB  (one processor)                                             \\
                                                & ranger - c++     & 0.94       & 7 mins           & --nthreads 20                                                     \\

@@

In order to asses these characteristics we ran \cursedforest\ classification on synthetics datasets with varying numbers of variables (features) and samples, similar to the dataset used in to evaluate ranger [REF: ranger], allocating varying number of CPU cores to the \cursedforest\ and also varying computational complexity of random forests by using range of \mtry\ values.

We investigate the different synthetic datasets generated for section 5.1. and measured the time of building a random forest model of 100 trees and the results reported below are averages of 5 runs. To improve consistency of measurements all the cases were executed with the same random seed.

Firstly we look at CurstedForest horizontal scalability for a medium size dataset of 2.5 million variables and 5000 samples, by varying \mtry\ fraction and the number of CPU cores allocated to the execution. The raw results are presented in the table [REF table1] and visualized in figure [REF: Fig 1] below
Regardless of the number of cores used, \cursedforest\ displays approximately linear dependency between the execution time and the \mtry\ parameter (presented here as a fraction of the number of variables) [Fig 1 a)

@@

\begin{table}[!ht]
\centering
\caption{
{\bf Time in seconds to build a random forest of 100
  trees with specified \mtry\ fraction using given number of CPU cores.}}
\begin{tabular}{|l|l|l|l|l|l|}
\hline
\multicolumn{2}{|l|}{\multirow{2}{*}{}}             & \multicolumn{4}{c|}{Number of CPU cores} \\
\cline{3-6}
\multicolumn{2}{|l|}{}                              & \bf{16} & \bf{32} & \bf{64} & \bf{128} \\
\hline
\multirow{4}{*}{\mtry\ (fraction)}        & \bf{0.10} & 708.3   & 340.0   & 186.0   & 134.5 \\
                                        & \bf{0.25} & 1134.9  & 535.0   & 270.8   & 179.6 \\
                                        & \bf{0.50} & 1724.1  & 856.6   & 428.7   & 243.8 \\
                                        & \bf{0.80} & 2313.5  & 1125.7  & 569.1   & 364.6 \\
\hline
\end{tabular}
\begin{flushleft} 
\end{flushleft}
\label{table10}
\end{table}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Feature Selection on real data 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ selects previously demonstrated cancer variants}
We consider the question of feature discovery on the Cancer Genome Atlas (TCGA) BRCA dataset.  We build this model on
VCF files retrieved from the TCGA data portal.  For the label, we use the cell type, tumor or normal.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 3: Classification selection
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ outperforms existing methods for classification problems}
In this section we compare \cursedforest 's performance against other published methods for predicting ethnicity on the
1000 Genomes Project dataset.

@@

\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.077 (XX) &                  &                                                                   \\
                                                & Spark ML  &            &                  &                                                                   \\
                                                & RF - R         & 0.05       & 24 min           & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes} \\
                                                & ranger - R      & 0.06       & 1 min 10 seconds & 16GB  (one processor)                                             \\
                                                & ranger - c++     & 0.06       & 7 mins           & --nthreads 20                                                     \\

@@

In order to asses these characteristics we ran \cursedforest\ classification on synthetics datasets with varying numbers
of variables (features) and samples, similar to the dataset used in to evaluate ranger \cite{Wright.and.Ziegle.2016},
allocating varying number of CPU cores to the \cursedforest\ and also varying computational complexity of random forests
by using a range of \mtry\ values.

We investigate the different synthetic datasets generated for section 5.1. and measured the time taken to build a random
forest model of 100 trees. The results reported below are averages of 5 runs, and all the cases were executed with the
same random seed, to improve consistency of measurements.

Firstly we look at CurstedForest horizontal scalability for a medium size dataset of 2.5 million variables and 5000
samples, by varying \mtry\ fraction and the number of CPU cores allocated to the execution. The raw results are
presented in table \ref{table10} and visualized in figure \ref{figure:synth} below. Regardless of the number of cores used,
\cursedforest\ displays approximately linear dependency between the execution time and the \mtry\ parameter (presented
here as a fraction of the number of variables) [Fig 1 a).

@@

\marginpar{can we remove this table and just rely on the figure?}
\begin{table}[!ht]
\centering
\caption{
{\bf Time in seconds to build a random forest of 100
  trees with specified \mtry\ fraction using given number of CPU cores.}}
\begin{tabular}{|l|l|l|l|l|l|}
\hline
\multicolumn{2}{|l|}{\multirow{2}{*}{}}             & \multicolumn{4}{c|}{Number of CPU cores} \\
\cline{3-6}
\multicolumn{2}{|l|}{}                              & \bf{16} & \bf{32} & \bf{64} & \bf{128} \\
\hline
\multirow{4}{*}{\mtry\ (fraction)}        & \bf{0.10} & 708.3   & 340.0   & 186.0   & 134.5 \\
                                        & \bf{0.25} & 1134.9  & 535.0   & 270.8   & 179.6 \\
                                        & \bf{0.50} & 1724.1  & 856.6   & 428.7   & 243.8 \\
                                        & \bf{0.80} & 2313.5  & 1125.7  & 569.1   & 364.6 \\
\hline
\end{tabular}
\begin{flushleft} 
\end{flushleft}
\label{table10}
\end{table}
